---
title: "project 1"
author: "Ehsan Pourhadi"
date: "July 31, 2018"
output: html_document
---

## Step no.1
##Loading and preprocessing the data


```{r, echo = TRUE}
##setting directory
activity<-read.csv("activity.csv")
```

##Some preliminaries

```{r}
dim(activity)
names(activity)
head(activity)
str(activity)
#total number of missing data
sum(is.na(activity$steps))/dim(activity)[[1]]
#converting the date column into date format using lubridate
library(lubridate)
activity$date<-ymd(activity$date)
length(unique(activity$date))
```

## Step no.2
##Histogram of the total number of steps taken each day

```{r, echo = TRUE}
library(ggplot2)
T1<-data.frame(tapply(activity$steps,activity$date,sum,na.rm=TRUE))
T1$date<-rownames(T1)
rownames(T1)<-NULL
names(T1)[[1]]<-"Total Steps"
png("plot1.png")
#Total Steps by date bar chart
ggplot(T1,aes(y=T1$`Total Steps`,x=T1$date))+geom_bar(stat="identity",color="red") + ylab("Total Steps")+xlab("Date")+ggtitle("Histogram of the total number of steps taken each day")
dev.off()
ggplot(T1,aes(y=T1$`Total Steps`,x=T1$date))+geom_bar(stat="identity", color="red") + ylab("Total Steps")+xlab("Date")+ggtitle("Histogram of the total number of steps taken each day")
#Histogram of total steps
qplot(T1$`Total Steps`,geom="histogram",xlab="Total Steps",ylab="Counts", main="Total Steps Historgram")
png("plot2.png")
qplot(T1$`Total Steps`,geom="histogram",xlab="Total Steps",ylab="Counts", main="Total Steps Historgram")
dev.off()
```
## Step no.3
##Mean and median number of steps taken each day

```{r, echo = TRUE}
library(dplyr)
T2<-data.frame(round(tapply(activity$steps,activity$date,mean,na.rm=TRUE),2))
T2$date<-rownames(T2)
rownames(T2)<-NULL
names(T2)[[1]]<-"Mean Steps"
temp<-activity%>%select(date,steps) %>% group_by(date) %>% summarise(median(steps))
names(temp)[[2]]<-"Median Steps"
T2$median<-temp$`Median Steps`
T2<-T2 %>% select(date,`Mean Steps`,median)
```

## Step no.4
##Time series plot of the average number of steps taken

```{r, echo = TRUE}
T3<-T2
T3$date<-as.Date(T3$date,format="%Y-%m-%d")
ggplot(T3,aes(x=T3$date,y=T3$`Mean Steps`))+geom_bar(stat="identity", color="blue")+scale_x_date()+ylab("Mean Steps Every day")+xlab("Date")+ggtitle("Mean Steps by Date")
png("plot3.png")
ggplot(T3,aes(x=T3$date,y=T3$`Mean Steps`))+geom_bar(stat="identity", color="blue")+scale_x_date()+ylab("Mean Steps Every day")+xlab("Date")+ggtitle("Mean Steps by Date")
dev.off()
```

## Step no.5
##The 5-minute interval that, on average, contains the maximum number of steps

```{r, echo = TRUE}
activity$interval<-factor(activity$interval)
T4<-aggregate(data=activity,steps~date+interval,FUN="mean")
T4<-aggregate(data=T4,steps~interval,FUN="max")
```

## Step no.6 
##Code to describe and show a strategy for imputing missing data

There are various techniques dealing with this question but only for the 
convenience we apply the mean/mode value substitution strategy to impute missing values.

```{r, echo = TRUE}
T5<-activity
T5$Missing<-is.na(T5$steps)
T5<-aggregate(data=T5,Missing~date+interval,FUN="sum")
T5.1<-data.frame(tapply(T5$Missing,T5$date,sum))
T5.1$date<-rownames(T5.1)
rownames(T5.1)<-NULL
names(T5.1)<-c("Missing","date")
T5.1$date<-as.Date(T5.1$date,format="%Y-%m-%d")

T5.2<-data.frame(tapply(T5$Missing,T5$interval,sum))
T5.2$date<-rownames(T5.2)
rownames(T5.2)<-NULL
names(T5.2)<-c("Missing","Interval")

par(mfrow=c(1,2))
plot(y=T5.1$Missing,x=T5.1$date,main="Missing Value Dist. by Date")
plot(y=T5.2$Missing,x=T5.2$Interval,main="Missing Value Dist. by Interval")
table(activity$date)
```

The plot shows that the missing values have a very disctinct pattern. For any interval, there always exist 8 missing values. For the date, there are consistantly 288 missing values. In aggregate, there are 8 dates that have missing value. We have no idea for the reason of these missing values but there is a pattern. So, we can find out the mean value imputation is suitable. 

Moreover, it is worth mentioning that every date has 288 data points. It means that the 8 dates have no data points. We can refine the analysis by take a look at these missing values depending on their Weekday and interval parameters to matach with the average.  


```{r, echo = TRUE}
#Dates that have missing values 
library(lubridate)
T5.3<-as.data.frame(T5.1) %>% select(date,Missing) %>% arrange(desc(Missing))
T5.3<-T5.3[which(T5.3$Missing!=0),]
T5.3$Weekday<-wday(T5.3$date,label=TRUE)
T5.4<-activity
T5.4$weekday<-wday(T5.4$date,label=TRUE)
#Finding the mean of steps every monday, and every interval
T5.5<-aggregate(data=T5.4,steps~interval+weekday,FUN="mean",na.rm=TRUE)
#Merge the pre-imputation table T5.4 table with the average table T5.5
T5.6<-merge(x=T5.4,y=T5.5,by.x=c("interval","weekday"),by.y=c("interval","weekday"),all.x=TRUE)
#Conditionally replacing the steps.x column NA value with the values from steps.y column value 
T5.6$Steps.Updated<-0
for (i in 1:dim(T5.6)[[1]]){
if(is.na(T5.6[i,3])){T5.6[i,6]=T5.6[i,5]}
else {T5.6[i,6]=T5.6[i,3]}
}
#Now simplify the imputed analytical data frame
T5.6 <-T5.6  %>% select(date,weekday,interval,Steps.Updated)
names(T5.6)[[4]]<-"Steps"

```
## Step no.7
## Histogram of the total number of steps taken each day after missing values are imputed

```{r, echo = TRUE}
png("plot4.png")
qplot(T5.6$Steps,geom="histogram",main="Histogram of the total number of steps taken each day after imputation",xlab="Steps",ylab="Count")
dev.off()
qplot(T5.6$Steps,geom="histogram",main="Histogram of the total number of steps taken each day after imputation",xlab="Steps",ylab="Count")

```

## Step no.8
##Panel plot comparing the average number of steps taken per 5-minute interval across weekdays and weekends

```{r, echo = TRUE}
install.packages("hexbin",repos = "http://cran.us.r-project.org")
T6<-T5.6
levels(T6$weekday)<-c(1,2,3,4,5,6,7)
T6$WDWE<-T6$weekday %in% c(1,2,3,4,5)
T6.1<-aggregate(data=T6,Steps~interval+WDWE,mean,na.rm=TRUE)
T6.1$WDWE<-as.factor(T6.1$WDWE)
levels(T6.1$WDWE)<-c("Weekend","Weekday")
png("plot5.png")
ggplot(data=T6.1,aes(y=Steps,x=interval,group=1,color=WDWE))+geom_line() +scale_x_discrete(breaks = seq(0, 2500, by = 300))+ylab("Mean Steps")+xlab("Intervals")+ggtitle("Mean steps across intervals by Weekend and Weekday")
dev.off()
ggplot(data=T6.1,aes(y=Steps,x=interval,group=1,color=WDWE))+geom_line() +scale_x_discrete(breaks = seq(0, 2500, by = 300))+ylab("Mean Steps")+xlab("Intervals")+ggtitle("Mean steps across intervals by Weekend and Weekday")

#Producing the panel plot
T6.1$interval<-as.numeric(as.character(T6.1$interval))
library(lattice)
xyplot(data=T6.1,Steps~interval|WDWE, grid = TRUE, type = c("p", "smooth"), lwd = 4,panel = panel.smoothScatter)
library(hexbin)
hexbinplot(data=T6.1,Steps~interval|WDWE, aspect = 1, bins=50)
png("plot6.png")
xyplot(data=T6.1,Steps~interval|WDWE, grid = TRUE, type = c("p", "smooth"), lwd = 4,panel = panel.smoothScatter)
dev.off()

png("plot7.png")
hexbinplot(data=T6.1,Steps~interval|WDWE, aspect = 1, bins=50)
dev.off()

```

